The Structure of $k$-Planar Graphs

### 2019-12-04 k-planar.tex
Beat this up into SoCG format

\documentclass{patmorin}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,wrapfig}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsfonts}
\usepackage{amsthm,mathtools}
\usepackage{pat}
\usepackage{paralist}
\usepackage{stmaryrd}
\usepackage{thm-restate}

@@

\usepackage[mathlines]{lineno}
% \setlength{\linenumbersep}{2em}
\linenumbers
% \rightlinenumbers
% \linenumbers
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
 \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
 \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
 \renewenvironment{#1}%
    {\linenomath\csname old#1\endcsname}%
    {\csname oldend#1\endcsname\endlinenomath}}% 
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
 \patchAmsMathEnvironmentForLineno{#1}%
 \patchAmsMathEnvironmentForLineno{#1*}}%
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
}

@@

 \newcommand{\note}[2]{{\color{red}[#1:~#2]}}
% \newcommand{\notex}[2]{{\color{blue}[#1:~#2]}}
\newcommand{\notex}[2]{}

@@

\title{\MakeUppercase{The Structure of $k$-Planar Graphs}}

\author{Vida Dujmovi\'c%
        \thanks{School of Computer Science and Electrical Engineering,
                University of Ottawa, Ottawa, Canada (\texttt{vida.dujmovic@uottawa.ca}).
                Research  supported by NSERC and the Ontario Ministry of Research and Innovation.},\,\,
        Pat Morin%
        \thanks{School of Computer Science, Carleton University, Ottawa, Canada (\texttt{morin@scs.carleton.ca}).                 Research  supported by NSERC and the Ontario Ministry of Research and Innovation.},\,\, and
        David R. Wood\thanks{School of Mathematics, Monash University, Melbourne, Australia (\texttt{david.wood@monash.edu}). Research supported by the Australian Research Council.}
}

\begin{document}
\begin{titlepage}
\maketitle

\begin{abstract}
Dujmovi\'c~et~al.~(FOCS 2019) recently proved that every planar graph is a subgraph of the strong product of a graph of bounded treewidth and a path. This tool has been used to solve longstanding problems on queue layouts, non-repetitive colouring, $p$-centered colouring, and implicit graph encoding. We extend this result for $k$-planar graphs, where a graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. In particular, we prove that every $k$-planar graph is a subgraph of the strong product of a graph of treewidth $O(k^5)$ and a path. This is the first result of this type for a non-minor-closed class of graphs. It implies, amongst other results, that $k$-planar graphs have non-repetitive chromatic number upper-bounded by a function of $k$. All these results generalise for drawings of graphs on arbitrary surfaces. In fact, we work in a much more general setting based on so-called shortcut systems that are of independent interest. 
\end{abstract}
\end{titlepage}

@@

A graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. Such graphs provide a natural generalisation of planar graphs, and are important in graph drawing research; see the recent bibliography on 1-planar graphs and the 140 references therein \citep{kobourov.liotta.ea:annotated}. The present paper studies the structure of $k$-planar graphs and other more general classes of graphs. 
\notex{PM}{I removed the following sentences because neither Vida nor could find anything about $k$-planar graphs in \cite{FP08} and, in any case, the proof is easy.  This could easily upset a referee.
% \note{PM}{I must be missing something.  The planarized graph $G'$ has $O(kn)$ vertices.  Let $S$ be the set of non-dummy vertices and take a separation $(A,B)$ of $G'$ with $|A\cap B|=O(\sqrt{kn})$, $|S\cap (A\setminus B)|\le 2|S|/3$ and $|S\cap (B\setminus A)|\le 2|S|/3$.  Now take $X=A\cap B$ plus the endpoints of every edge of $G$ that contains a dummy vertex in $A\cap B$.  $|X|\le 5|A\cap B|=O(\sqrt{kn})$ and $X$ is a separator of $G$.} \note{DW}{I think you have just found a new proof of this fact. Note that you are actually using that $G'$ has treewidth $O(\sqrt{kn})$ and then applying the Robertson-Seymour separator lemma for graphs of given treewidth (so that $|S\cap (A\setminus B)|\le 2|S|/3$ and $|S\cap (B\setminus A)|\le 2|S|/3$). }
Various structural results about planar graphs generalise for $k$-planar graphs. For example, \citet{FP08} generalised the Lipton-Tarjan separator theorem to show that every $k$-planar graph with $n$ vertices has a balanced separator of order $O(\sqrt{kn})$.}

@@

\notex{DW}{Is some sort of converse to \cref{generalized-tripod}  also true? \cref{generalized-tripod}  shows that if a graph $G$ has a layered partition with all the nice properties, then $G$ has this generalised tripod partition. Isn't also true that if $G$ has a generalised tripod partition, then $G$ has a layered partition with all the nice properties. The proof is basically in \citep{dujmovic.joret.ea:planar}, right?}
\notex{PM}{Yes, if by generalized tripod partition you mean $T$ and $\mathcal{Y}=(Y_x:x\in V(T))$ that satisify \yref{thickness}, \yref{separator}, and \yref{ancestor-edge}.  From \yref{separator} and \yref{ancestor-edge} you can construct a width-$t$ $T$-decomposition of $H=G/\mathcal{Y}$.  By \yref{thickness}, $\mathcal{Y}$, has layered width at most $\ell$.}

@@

The following converse of \cref{gAddDummyDegree} provides a sufficient condition for a graph to be $(g,k)$-planar that is slightly more restrictive than being obtained from a $(k,d)$-shortcut system of a $(g,k)$-planar graph. \notex{DW}{Is this sentence clear enough?}\notex{PM}{Sure.}

>>>>

\documentclass[a4paper,UKenglish,cleveref, autoref]{lipics-v2019}

% \documentclass{socg-lipics-v2019}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,wrapfig}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsfonts}
\usepackage{amsthm,mathtools}
\usepackage{pat}
\usepackage{paralist}
\usepackage{stmaryrd}
\usepackage{thm-restate}

@@

% \usepackage[mathlines]{lineno}
% % \setlength{\linenumbersep}{2em}
% \linenumbers
% % \rightlinenumbers
% % \linenumbers
% \newcommand*\patchAmsMathEnvironmentForLineno[1]{%
%  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
%  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
%  \renewenvironment{#1}%
%     {\linenomath\csname old#1\endcsname}%
%     {\csname oldend#1\endcsname\endlinenomath}}% 
% \newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
%  \patchAmsMathEnvironmentForLineno{#1}%
%  \patchAmsMathEnvironmentForLineno{#1*}}%
% \AtBeginDocument{%
% \patchBothAmsMathEnvironmentsForLineno{equation}%
% \patchBothAmsMathEnvironmentsForLineno{align}%
% \patchBothAmsMathEnvironmentsForLineno{flalign}%
% \patchBothAmsMathEnvironmentsForLineno{alignat}%
% \patchBothAmsMathEnvironmentsForLineno{gather}%
% \patchBothAmsMathEnvironmentsForLineno{multline}%
% }

@@

 % \newcommand{\note}[2]{{\color{red}[#1:~#2]}}
% \newcommand{\notex}[2]{{\color{blue}[#1:~#2]}}
% \newcommand{\notex}[2]{}

@@

% \title{\MakeUppercase{The Structure of $k$-Planar Graphs}}
\title{The Structure of $k$-Planar Graphs}
\titlerunning{The Structure of $k$-Planar Graphs}

\author{Vida Dujmović}{School of Computer Science and Electrical Engineering, University of Ottawa}{vida.dujmovic@uottawa.ca}{[0000-0003-0471-4118]}{Research  supported by NSERC and the Ontario Ministry of Research and Innovation.}
% , Ottawa, Canada (\texttt{vida.dujmovic@uottawa.ca}).
%                 Research  supported by NSERC and the Ontario Ministry of Research and Innovation.},\,\,
\author{Pat Morin}{School of Computer Science, Carleton University}{morin@scs.carleton.ca}{[0000-0003-0471-4118]}{Research supported by NSERC.}
% , Ottawa, Canada (\texttt{morin@scs.carleton.ca}).                 Research  supported by NSERC and the Ontario Ministry of Research and Innovation.},\,\, and
\author{David R. Wood}{School of Mathematics, Monash University}{david.wood@monash.edu}{[0000-0003-0471-4118]}{Research supported by the Australian Research Council.}
% , Melbourne, Australia (\texttt{david.wood@monash.edu}). Research supported by the Australian Research Council.}
% }

\authorrunning{V. Dujmović, P. Morin, and D. R. Wood}

\keywords{$k$-planar graphs, layered partitions, queue-number, nonrepetitive colouring, centered colouring, $k$-nearest-neighbour graphs}

\Copyright{Vida Dujmović, Pat Morin, and David R. Wood}

\begin{document}
% \begin{titlepage}
\maketitle

\begin{abstract}
Dujmovi\'c~et~al.~(FOCS 2019) recently proved that every planar graph is a subgraph of the strong product of a graph of bounded treewidth and a path. This tool has been used to solve longstanding problems on queue layouts, non-repetitive colouring, $p$-centered colouring, and implicit graph encoding. We extend this result for $k$-planar graphs, where a graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. In particular, we prove that every $k$-planar graph is a subgraph of the strong product of a graph of treewidth $O(k^5)$ and a path. This is the first result of this type for a non-minor-closed class of graphs. It implies, amongst other results, that $k$-planar graphs have non-repetitive chromatic number upper-bounded by a function of $k$. All these results generalise for drawings of graphs on arbitrary surfaces. In fact, we work in a much more general setting based on so-called shortcut systems that are of independent interest. 
\end{abstract}
% \end{titlepage}

@@

A graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. Such graphs provide a natural generalisation of planar graphs, and are important in graph drawing research; see the recent bibliography on 1-planar graphs and the 140 references therein \citep{kobourov.liotta.ea:annotated}. The present paper studies the structure of $k$-planar graphs and other more general classes of graphs. 
% \notex{PM}{I removed the following sentences because neither Vida nor could find anything about $k$-planar graphs in \cite{FP08} and, in any case, the proof is easy.  This could easily upset a referee.
% % \note{PM}{I must be missing something.  The planarized graph $G'$ has $O(kn)$ vertices.  Let $S$ be the set of non-dummy vertices and take a separation $(A,B)$ of $G'$ with $|A\cap B|=O(\sqrt{kn})$, $|S\cap (A\setminus B)|\le 2|S|/3$ and $|S\cap (B\setminus A)|\le 2|S|/3$.  Now take $X=A\cap B$ plus the endpoints of every edge of $G$ that contains a dummy vertex in $A\cap B$.  $|X|\le 5|A\cap B|=O(\sqrt{kn})$ and $X$ is a separator of $G$.} \note{DW}{I think you have just found a new proof of this fact. Note that you are actually using that $G'$ has treewidth $O(\sqrt{kn})$ and then applying the Robertson-Seymour separator lemma for graphs of given treewidth (so that $|S\cap (A\setminus B)|\le 2|S|/3$ and $|S\cap (B\setminus A)|\le 2|S|/3$). }
% Various structural results about planar graphs generalise for $k$-planar graphs. For example, \citet{FP08} generalised the Lipton-Tarjan separator theorem to show that every $k$-planar graph with $n$ vertices has a balanced separator of order $O(\sqrt{kn})$.}

@@

% \notex{DW}{Is some sort of converse to \cref{generalized-tripod}  also true? \cref{generalized-tripod}  shows that if a graph $G$ has a layered partition with all the nice properties, then $G$ has this generalised tripod partition. Isn't also true that if $G$ has a generalised tripod partition, then $G$ has a layered partition with all the nice properties. The proof is basically in \citep{dujmovic.joret.ea:planar}, right?}
% \notex{PM}{Yes, if by generalized tripod partition you mean $T$ and $\mathcal{Y}=(Y_x:x\in V(T))$ that satisify \yref{thickness}, \yref{separator}, and \yref{ancestor-edge}.  From \yref{separator} and \yref{ancestor-edge} you can construct a width-$t$ $T$-decomposition of $H=G/\mathcal{Y}$.  By \yref{thickness}, $\mathcal{Y}$, has layered width at most $\ell$.}

@@

The following converse of \cref{gAddDummyDegree} provides a sufficient condition for a graph to be $(g,k)$-planar that is slightly more restrictive than being obtained from a $(k,d)$-shortcut system of a $(g,k)$-planar graph.
